---
title: "Analysis of the association between fluid balance and mortality accounting for changes in disease severity"
author: "S.G. Blok & H.J. de Grooth"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    number_sections: true
    toc_float: true
    code_folding: hide
---

# Load libraries and data

```{r setup, warning=FALSE}
library(dplyr)
library(kableExtra)
library(tibble)
library(survival)
library(survminer)
library(ggplot2)
library(cmprsk)
library(survey)

if (rstudioapi::isAvailable()) {
  # Get the path of the current Rmd file in RStudio
  current_path <- dirname(rstudioapi::getActiveDocumentContext()$path)
} else {
  # Alternative method for non-RStudio environments
  current_path <- dirname(normalizePath(knitr::current_input()))
}

# Set the working directory
setwd(current_path)

# Load the .rds files
d_daily <- readRDS("d_daily.rds")
d <- readRDS("d.rds")

```

## Filter the data

```{r}
d_daily <- d_daily %>%
  filter(day > 1)
```

# Logistic regression

We will first use a logistic regression model to investigate the association with fluid balance and mortality at day 90

## Logistic regression using cumulative fluid balance

```{r}
# Ensure that mort_90days is a binary factor
d$mort_90days <- as.factor(d$mort_90days)

# Fit the logistic regression model
cfb_logit_mod <- glm(mort_90days ~ fluids_cumulative, data = d, family = binomial)

# Extract and display the model summary in a cleaner format
model_summary <- summary(cfb_logit_mod)
knitr::kable(as.data.frame(coef(model_summary)), 
             caption = "Logistic Regression Model Coefficients")

# Calculate odds ratios and 95% confidence intervals
odds_ratios_cum <- exp(cbind(OR = coef(cfb_logit_mod), confint(cfb_logit_mod)))

# Display the odds ratios
knitr::kable(odds_ratios_cum, digits = 3, caption = "Odds Ratios and 95% Confidence Intervals")
```

## Using mean fluid balance

Of course the previous results are confounded by length of stay thereby giving the false impression that cumulative fluid balance is protective for mortality. Perhaps we should better use the mean fluids (defined as fluids_cumulative/los)

```{r}
# Fit the logistic regression model
mfb_logit_mod <- glm(mort_90days ~ fluids_mean, data = d, family = binomial)

# Extract and display the model summary in a cleaner format
model_summary <- summary(mfb_logit_mod)
knitr::kable(as.data.frame(coef(model_summary)), 
             caption = "Logistic Regression Model Coefficients")

# Calculate odds ratios and 95% confidence intervals
odds_ratios_mean <- exp(cbind(OR = coef(mfb_logit_mod), confint(mfb_logit_mod)))

# Display the odds ratios
knitr::kable(odds_ratios_mean, digits = 3, caption = "Odds Ratios and 95% Confidence Intervals")
```

## Correcting for baseline disease severity

Now we get a much different idea. Mean fluid balance is strongly associated with mortality in this model. Commonly, we then correct for severity of disease at baseline using disease severity scores. Lets do so.

```{r}
# Fit the logistic regression model
sev_logit_mod <- glm(mort_90days ~ fluids_mean + severity_baseline, data = d, family = binomial)


# Extract and display the model summary in a cleaner format
model_summary <- summary(sev_logit_mod)
knitr::kable(as.data.frame(coef(model_summary)), 
             caption = "Logistic Regression Model Coefficients")

# Calculate odds ratios and 95% confidence intervals
odds_ratios_sev <- exp(cbind(OR = coef(sev_logit_mod), confint(sev_logit_mod)))

# Display the odds ratios
knitr::kable(odds_ratios_sev, digits = 3, caption = "Odds Ratios and 95% Confidence Intervals")
```

## Summary

```{r}


# Define rows and assign values manually
final_table <- data.frame(
  Variable = c(
    "Cumulative Fluid Balance", "fluids_cumulative",
    "Mean Fluid Balance", "fluids_mean",
    "Mean Fluid Balance Adjusted", "fluids_mean", "severity_baseline"
  ),
  `Odds Ratio` = c(
    "", 0.953,
    "", 5.479,
    "", 4.605, 1.182
  ),
  `95% CI Lower` = c(
    "", 0.923,
    "", 3.654,
    "", 2.993, 1.020
  ),
  `95% CI Upper` = c(
    "", 0.983,
    "", 8.339,
    "", 7.184, 1.371
  ),
  stringsAsFactors = FALSE
)

# Format the table
final_table %>%
  kable(
    format = "html",
    align = "lccc",
    col.names = c("Variable", "Odds Ratio", "95% CI Lower", "95% CI Upper"),
    caption = "Overview of Odds Ratios and 95% Confidence Intervals"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE
  ) %>%
  row_spec(c(1, 3, 5), bold = TRUE, background = "#f2f2f2")  # Highlight category rows
```

When we work with non-simulated data, we usually consider more covariates such as age or admission diagnosis.

# Time dependent Cox model

These type of associations are sometimes also analyzed with a time dependent Cox model (<https://link.springer.com/article/10.1186/s13054-015-0970-1>[)](https://link.springer.com/article/10.1186/s13054-015-0970-1)

We will also construct such a model.\
\
Please be aware that we have only 10 days of survival data available here and we have fluids data for every day the patient was alive and in the ICU.\

## First, plot mortality in the first 10 days

```{r}
# Filter out rows where both death and discharge are NA
filtered_data <- d_daily %>%
  filter(!(is.na(death) & is.na(discharge)))

# Collapse data to one row per patient
km_data <- filtered_data %>%
  group_by(pt) %>%
  summarise(
    time = max(day, na.rm = TRUE),         # Time to event or censoring
    status = ifelse(any(death == 1), 1, 0) # Event status: 1 if death occurred, 0 otherwise
  ) %>%
  ungroup()

# Fit Kaplan-Meier survival curve
km_fit <- survfit(Surv(time, status) ~ 1, data = km_data)

# Plot Kaplan-Meier curve
ggsurvplot(
  km_fit,
  data = km_data,
  conf.int = TRUE,                # Add confidence intervals
  risk.table = TRUE,              # Show number at risk below the plot
  ggtheme = theme_minimal(),      # Use a minimal theme
  title = "Kaplan-Meier Curve for Mortality",
  xlab = "Time (Days)",
  ylab = "Survival Probability"
)
```

## Cox proportional hazards regression using total fluids

```{r}

# Filter out rows where both death and discharge are NA
filtered_data <- d_daily %>%
  filter(!(is.na(death) & is.na(discharge)))

# Collapse data to one row per patient for survival analysis
cox_data <- filtered_data %>%
  group_by(pt) %>%
  summarise(
    time = max(day, na.rm = TRUE),         # Time to event or censoring
    status = ifelse(any(death == 1), 1, 0), # Event indicator: 1 if death occurred, 0 otherwise
    fluids = sum(fluids, na.rm = TRUE)     # Total fluids as a summary measure (or choose another)
  ) %>%
  ungroup()

# Fit the Cox model with fluids as a covariate
cox_model <- coxph(Surv(time, status) ~ fluids, data = cox_data)

# Summary of the model
summary(cox_model)
```

This of course is again quite biased by LOS. Cumulative fluid balance is now again protective for death as the patients that do not die can keep accumulating fluids.

## Cox proportional hazards with time-varying fluid as a covariate

```{r}

cox_time_varying_data <- d_daily %>%
  filter(!(is.na(death) & is.na(discharge))) %>%  # Remove irrelevant rows
  group_by(pt) %>%
  mutate(
    last_day = max(day, na.rm = TRUE),            # Patient's last observed day
    event_day = ifelse(death == 1, day, NA),      # Day of death if applicable
    status = ifelse(death == 1 & day == last_day, 1, 0)  # Status = 1 for death interval
  ) %>%
  ungroup() %>%
  mutate(
    start = day - 1,  # Start of interval
    stop = day        # Stop of interval
  )

# Fit the Cox model with time-varying covariates
cox_time_varying_model <- coxph(
  Surv(start, stop, status) ~ fluids,
  data = cox_time_varying_data
)

# Summary of the model
summary(cox_time_varying_model)
```

## Cox proportional hazards with time-varying fluid and baseline disease severity score as a covariate

```{r}
# Step 1: Extract the first measured severity for each patient
first_severity <- d_daily %>%
  group_by(pt) %>%
  summarise(severity_first = first(severity[!is.na(severity)])) %>% # First non-NA severity
  ungroup()

# Step 2: Merge the static first severity into the time-varying dataset
cox_static_severity_data <- cox_time_varying_data %>%
  left_join(first_severity, by = "pt")             # Add static severity

# Step 3: Fit the Cox model with static severity
cox_model_static_severity <- coxph(
  Surv(start, stop, status) ~ fluids + severity_first,
  data = cox_static_severity_data
)

# Summary of the model
summary(cox_model_static_severity)
```

## Cox proportional hazards with time-varying fluid and time varying disease severity score as a covariate

```{r}
cox_model_dynamic_severity <- coxph(
  Surv(start, stop, status) ~ fluids + severity,
  data = cox_time_varying_data
)

# Summary of the adjusted model
summary(cox_model_dynamic_severity)
```

## Summary

```{r}
# Extract key results from each Cox model
model_results <- data.frame(
  Model = c(
    "Unadjusted (fluids only)",
    "Time-Varying Covariate (fluids)",
    "Baseline Severity Adjusted (fluids)",
    "Baseline Severity Adjusted (severity)",
    "Dynamic Severity Adjusted (fluids)",
    "Dynamic Severity Adjusted (severity)"
  ),
  HR = c(
    exp(coef(cox_model)["fluids"]),
    exp(coef(cox_time_varying_model)["fluids"]),
    exp(coef(cox_model_static_severity)["fluids"]),
    exp(coef(cox_model_static_severity)["severity_first"]),
    exp(coef(cox_model_dynamic_severity)["fluids"]),
    exp(coef(cox_model_dynamic_severity)["severity"])
  ),
  CI_Lower = c(  # Changed to use _
    summary(cox_model)$conf.int["fluids", "lower .95"],
    summary(cox_time_varying_model)$conf.int["fluids", "lower .95"],
    summary(cox_model_static_severity)$conf.int["fluids", "lower .95"],
    summary(cox_model_static_severity)$conf.int["severity_first", "lower .95"],
    summary(cox_model_dynamic_severity)$conf.int["fluids", "lower .95"],
    summary(cox_model_dynamic_severity)$conf.int["severity", "lower .95"]
  ),
  CI_Upper = c(  # Changed to use _
    summary(cox_model)$conf.int["fluids", "upper .95"],
    summary(cox_time_varying_model)$conf.int["fluids", "upper .95"],
    summary(cox_model_static_severity)$conf.int["fluids", "upper .95"],
    summary(cox_model_static_severity)$conf.int["severity_first", "upper .95"],
    summary(cox_model_dynamic_severity)$conf.int["fluids", "upper .95"],
    summary(cox_model_dynamic_severity)$conf.int["severity", "upper .95"]
  ),
  P_value = c(
    summary(cox_model)$coefficients["fluids", "Pr(>|z|)"],
    summary(cox_time_varying_model)$coefficients["fluids", "Pr(>|z|)"],
    summary(cox_model_static_severity)$coefficients["fluids", "Pr(>|z|)"],
    summary(cox_model_static_severity)$coefficients["severity_first", "Pr(>|z|)"],
    summary(cox_model_dynamic_severity)$coefficients["fluids", "Pr(>|z|)"],
    summary(cox_model_dynamic_severity)$coefficients["severity", "Pr(>|z|)"]
  ),
  stringsAsFactors = FALSE
)

# Round values for readability
model_results <- model_results %>%
  mutate(
    HR = round(HR, 2),
    CI_Lower = round(CI_Lower, 2),
    CI_Upper = round(CI_Upper, 2),
    P_value = formatC(P_value, format = "e", digits = 2)
  )

# Create a neat summary table
model_results %>%
  kbl(
    col.names = c("Model", "Hazard Ratio (HR)", "CI Lower", "CI Upper", "P-value"),
    caption = "Summary of Cox Models for Fluids and Severity"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE
  )
```

# Taking competing risks into account

One important source of possible bias in our previous analysis is that patients are censored at ICU discharge. One way to adjust for this is to use a Fine-Gray Subdistribution Hazard model. Lets do this

## Fine-Gray competing risk analysis for mean fluid balance unadjusted

```{r}

mean_fluids <- d %>%
  select(pt, fluids_mean)

cr_data <- d_daily %>%
  filter(!(is.na(death) & is.na(discharge))) %>%  # Remove rows with no outcomes
  group_by(pt) %>%
  summarise(
    time = max(day, na.rm = TRUE),         # Time to event or censoring
    event = case_when(
      any(death == 1) ~ 1,                 # Event = 1 for death
      any(discharge == 1) ~ 2,             # Event = 2 for discharge
      TRUE ~ 0                             # Event = 0 for censoring
    )
  ) %>%
  ungroup() %>%
    left_join(mean_fluids, by = "pt")  # Add mean fluids to the competing risks dataset





library(cmprsk)

# Fine-Gray model for death (event = 1, competing event = 2) with fluids
fg_unadjusted <- crr(
  ftime = cr_data$time,         # Time-to-event
  fstatus = cr_data$event,      # Event type (1 = death, 2 = discharge, 0 = censored)
  cov1 = as.matrix(cr_data$fluids_mean)  # Fluids as a covariate
)

# Summary of the Fine-Gray model
summary(fg_unadjusted)
```

## Fine-Gray competing risk analysis with adjustment for baseline disease severity

```{r}
mean_fluids_severity_baseline <- d %>%
  select(pt, fluids_mean, severity_baseline)

cr_data <- d_daily %>%
  filter(!(is.na(death) & is.na(discharge))) %>%  # Remove rows with no outcomes
  group_by(pt) %>%
  summarise(
    time = max(day, na.rm = TRUE),         # Time to event or censoring
    event = case_when(
      any(death == 1) ~ 1,                 # Event = 1 for death
      any(discharge == 1) ~ 2,             # Event = 2 for discharge
      TRUE ~ 0                             # Event = 0 for censoring
    )
  ) %>%
  ungroup() %>%
    left_join(mean_fluids_severity_baseline, by = "pt")  # Add mean fluids to the competing risks dataset





library(cmprsk)

# Fine-Gray model for death (event = 1, competing event = 2) with fluids
fg_adjusted <- crr(
  ftime = cr_data$time,         # Time-to-event
  fstatus = cr_data$event,      # Event type (1 = death, 2 = discharge, 0 = censored)
  cov1 = as.matrix(cr_data[, c("fluids_mean", "severity_baseline")])  # Fluids as a covariate
)

# Summary of the Fine-Gray model
summary(fg_adjusted)
```

So once again we see that if we look at mean fluid balance with or without adjustment for baseline disease severity, there will be an apparent association. Previously, we could take into account varying disease severity. We could do this with landmark analysis, but we lose granularity. We could also try subdistribution hazards with Cox models and time varying disease severity (like previously). However, in that case we would assume death and ICU discharge are indepedent events which is clearly not the case.

# Marginal structural modeling

One solution could be marginal structural modeling. Lets try.

```{r}

# Prepare the long-format dataset
msm_data <- d_daily %>%
  filter(!(is.na(death) & is.na(discharge))) %>%
  mutate(
    start = day - 1,
    stop = day,
    event = case_when(
      death == 1 ~ 1,       # Death as event of interest
      discharge == 1 ~ 2,   # Discharge as competing risk
      TRUE ~ 0              # Censored
    )
  )

# Fit a logistic regression model for stabilized weights
#ipw_model <- glm(
#  fluids ~ severity + start,  # Predict fluids based on severity and time
#  data = msm_data,
#  family = binomial()
#)


```
